Here I am, having to do the impossible: acknowledge everyone who helped me through the six years it took me to complete this PhD dissertation. I'll do my best and hope not to forget anyone. I have to start by saying that doing this PhD is one of the biggest dreams of my life. And I am thrilled that I've managed to make it a reality. I wouldn't even be able to start this if it wasn't for my parents, who always supported my dreams, providing everything they could to help me succeed. After my parents, I want to thank my siblings Lorrana and Leonardo for their love and support.

I also would like to express my deepest gratitude to my PhD advisor Jim Whitehead, who gave me a lot of freedom to explore my research interests while guiding my work with enthusiasm, patience, and high standards. I'm also extremely grateful to Levi Lelis, who participated in this work since the beginning, helping me with countless hours of very productive discussions. Throughout my time at UC Santa Cruz, I had the chance to work with other great professors. In particular, I want to thank James Davis, who I worked for as a TA for more than two years. All this time working with James was fundamental for me to develop my own style of teaching.

My friends are also essential for the success of this research. First, I want to thank Julia for both the support during my Masters and during the PhD application process. The latter was a tough time in which she very kindly adjusted her life to stay by my side as much as possible. Second, I want to thank  Alessio, Caio, Henrique, and Vinicius, some of my best friends, who traveled from Brazil to Santa Cruz to visit me. Third, I want to thank all the friends I have made in Santa Cruz: Chandranil, Bia, Gabriel, Ivo, Nalini, Vishakha, Morteza, Keshav, Ankit, Abhinav, Yona, Vini, and Gen. Fourth, I want to thank Juninho, Jo\~{a}o, Yasmin, Caio (again), and Oliver. Besides being amazing friends since I joined college, they have been sharing daily (remote) moments with me since the COVID-19 pandemic started, which have been vital for me to pass through this pandemic with sanity. I want to especially thank Jo\~{a}o and Yasmin for helping me making the figures for this dissertation. Last but not least, I want to thank Gabi for being by my side in the very last eight months of my life, which have been extremely busy with the work to finish this dissertation.

I'm also grateful to the Conselho Nacional de Pesquisa (CNPq), the Brazilian agency that funded the first four years of this work. I wouldn't have been able to develop this research without this support.
